---
title: "DATA-413/613 HW 3: Lists and Iteration"
author: "Ana Miranda"
number-sections: true
toc: false
format:
  html
---

# Working with Lists

Because of their ability to handle general non-rectangular data, lists (or list-like objects) are often the output of procedures in R or from the collection of complex data. 

- The file `discog_masters_100.rds` contains output from a query of the [Discog](https://www.discogs.com/) database. The data are about 100 releases of "Masters" albums in the US in 2023.

1.  Use `readr::read_rds()` and a relative path to read this data from the data folder or `readRDS(url("https://raw.githubusercontent.com/AU-datascience/data/main/413-613/discog_masters_100.rds", open="rb"))` to read from GitHub.

- If you use the latter, explain what the `url() argument `rb` is doing.

```{r}

library(tidyverse)

```


```{r}

discog_masters_100 <- readRDS("../data/discog_masters_100.rds")

```


2.  Use appropriate functions to obtain the length of the list and then the vector of the names of the elements in the list.

```{r}

length(discog_masters_100)
names(discog_masters_100)

```

3.  The `pagination` element does not provide any information of interest. Remove it from the list in one step.

```{r}

discog_masters_100$pagination <- NULL
names(discog_masters_100)

```

4.  There should only be one element in the list, `results` which is a list of 100 elements - each element is the data for one masters album. 

```{r}

length(discog_masters_100)
length(discog_masters_100$results)

```

a. To simplify working with the data, assign the `results` element back to `masters` so `masters` is now a list with 100 elements. 

```{r}

#names(discog_masters_100)[1] <- "masters"

masters <- discog_masters_100$results
class(masters)
length(masters)

```


b. Use a function to show just the number of elements in the list.

```{r}

length(masters)

```

c. Use a {purrr} function to generate a vector with the number of elements in each of the top level elements in the list.

```{r}

map_int(masters, length)

```

5.  Extract elements from a list.

a.  Extract the first element from `masters` so it has one element.

```{r}

first_elem <- masters[1]
str(first_elem)

```

b.  Extract the first element from `masters` so it has 19 elements.

```{r}

first_elem_list19 <- masters[[1]]
str(first_elem_list19)

```

c.  Extract the first element of the first album of `masters` so it is a list.

```{r}

first_elem_first_album <- masters[[1]][1]
str(first_elem_first_album)

```

d.  Extract the first element of the first album of `masters` so it is a character.

```{r}

first_album <- masters[[1]][[1]]
str(first_album)

```

e.  Use `purrr::pluck` to get the last album in `masters` in a way where it does not matter if the length of `masters` changes.

```{r}

last_album <- pluck(masters, length(masters))
str(last_album)

```

f.  Use `purrr::pluck` to get the last element of the last album in `masters`.

```{r}

last_elem_last_album <- pluck(masters, length(masters), length(pluck(masters, length(masters))))
str(last_elem_last_album)

```

g. Show two ways to get the `title` of the second album in `masters`.

```{r}

sec_album_title1 <- masters[[2]]$title
sec_album_title1

sec_album_title2 <- pluck(masters, 2, "title")
sec_album_title2

```

6.  Operating on elements within a list.

a.  Use a {purrr} function to create a vector with the class of elements in the list.

```{r}

class_100_elem_masters <- map_chr(masters, ~ paste(class(.x), collapse = ", "))
class_100_elem_masters

```

b.  Use {purrr} functions to create a list of length 100, where each element of the list is a named vector with the class of each of the 19 elements in one of the 100 list elements in `masters`. **Show just the first three elements**.

```{r}

class_list <- map(masters, function(album) {
  classes <- map_chr(album, typeof)
  setNames(unlist(classes), names(album))
})
#class_list

head(class_list, n=3)

```


c. Use a {purrr} function to create a list of length 100 where each element is a character vector with the names of 19 elements in one of the 100 list elements in `masters`.**Show just the last three elements**.

```{r}

names_list <- map(masters, names)
tail(names_list, n=3)

```


d. It appears that each of the 100 list elements in `masters` is a list of 19 elements for each album where the elements have the same classes and names across all albums. To confirm this using code, not just visual inspection, use a {purrr} function to create a list of 100 tibbles where each tibble contains the names of each of the 19 elements and then convert that list into a tibble with 1900 rows and extract the distinct names and sort them in alphabetical order. You should have 19 names.

```{r}

#map(masters, names)

df_names <- map(masters, ~ tibble(names = names(.x)))
combined_df <- list_rbind(df_names)

combined_df %>%
  distinct(names) %>%
  arrange(names)

```


e. Extract the list of the titles from each Album and convert to a tibble with column name `album_name` and show the last six entries.

```{r}

titles_df <- tibble(album_name = map_chr(masters, ~ .x$title))
tail(titles_df, n=6)

```


f. Run the following code and explain what it is doing.

```{r}

map(masters, \(ls) {
  ls$community <- as_tibble(ls$community)
  ls
}) -> masters2
#masters2

```

** Answer:** This code seems to interact through the entire list of masters and returns its all values back to masters, creating an exact copy of masters.



# For Loops

Consider the recursive sequence defined by:

$$
x_n = x_{n-1} + \frac{|x_{n-2} - x_{n-3}|}{5}.
$$

In words, element $x_n$ is the sum of element $x_{n-1}$ and the absolute value of the difference between between elements $x_{n-2}$ and $x_{n-3}$, divided by five. For example, if we let $x_1 = 3$, $x_2 = 1$, and $x_3 = 10$, then $x_4$ is

$$
x_4 = 10 + \frac{|1 - 3|}{5} = 10.4.
$$

1.  Write a function called `calcn()` to calculate $x_n$ for any positive $n$ based on input arguments of a vector `x` with the first three elements of this sequence ($x_1$, $x_2$, $x_3$) and a a second input argument of an integer `n` denoting the final element of the sequence to calculate.

-   `calcn(x, n)` should return the value of the `n`^th^ element of the sequence.
-   Include error checking to ensure both inputs are of the **correct length and class** and **`n` is greater than 0**.

For example, in my implementation of `calcn()`, I obtained the following:

```{r}

calcn <- function(x, n) {
  if (length(x) < 3) {
    stop("Error: The x vector must have at least three elements.")
  }
  
  if (n <= 3) {
    return(paste0("The x", n, " element of the sequence is: ", x[n]))
  }
  
  x_n <- seq_along(n)
  x_n[1:3] <- x
  
  for (i in 4:n) {
    x_n[i] <- x_n[i - 1] + abs(x_n[i - 2] - x_n[i - 3]) / 5
  }
  
  return(paste0("The x", n, "th element of the sequence is: ", x_n[n]))
}

calcn(x = c(2, 4, 3), n = 3L)
calcn(x = c(2, 4, 3), n = 4L)
calcn(x = c(2, 4, 3), n = 5L)
calcn(x = c(2, 4, 3), n = 6L)
calcn(x = c(2, 4, 3), n = 7L)

```

- Evaluate your function at the following inputs:

```{r}

calcn(c(11, 1, 130), 1000L)
calcn(c(11, 1, 130), 1L)
calcn(c(7, 3, 20), 8L)

```


# Question: Lists, For-loops, and `map_*()`

Lists are often used to save simulation output. You can then extract individual elements from the lists using for-loops.

Consider the $t$-test, often used to test whether or not the true mean of a population for some observations is 0. We can use the following code to simulate data from a [Normal ($\mu = 0$,$\sigma = 1$) distribution](https://en.wikipedia.org/wiki/Normal_distribution), and then use a $t$-test to test if the true mean is 0:

```{r}

set.seed(1234)
x <- rnorm(n = 10, mean = 0, sd = 1)
x

tout <- t.test(x)
tout

```

1.  The output of `t.test()` is a list. Use one function to show how many elements are in the top level of the list along with their names and class.

```{r}

str(tout)

```


2.  Write a for-loop to perform the following three operations on iteration `i`:
    1.  Draw 12 random observations from a Normal distribution with mean 0 and standard deviation 3.
    2.  Run a $t$-test on these 12 observations.
    3.  **Save the output of the** $t$-test as the `i`^th^ element in a list called `tlist`.

-   Use `set.seed()` to set the seed to 1234.
-   Run for 1000 iterations.

```{r}

set.seed(1234)
tlist <- vector(mode = "list", length = 1000)

for (i in seq_along(tlist)) {

    x <- rnorm(n = 12, mean = 0, sd = 3)
    
    tlist[[i]] <- t.test(x)
}

tlist[[1]]
tlist[[1000]]

```


3.  Use the appropriate `purrr:map_*` function to extract the sample mean from each test in `tlist` (from 2) and pipe to an appropriate plot to show the sampling distribution of the sample mean.

-   Hint: Make sure the data going into `ggplot()` is a data frame (tibble).
-   Look at help for `tibble` for `.name_repair =` to create a useful name for the variable.

```{r}

sample_means <- map_dbl(tlist, "estimate")
str(sample_means)

means_df <- tibble(sample_mean = sample_means)
glimpse(means_df)

ggplot(means_df, aes(x = sample_mean)) +
  geom_histogram(fill = "gray", color = "black", alpha = 0.5) +
  
  theme_bw() +

  xlab("Sample Mean") +
  ylab("Frequency") +
  ggtitle(label = "Sampling Distribution of the Sample Mean from 1000 Normal Random Samples",
          subtitle = "(mean=0 and sd=3)") +
    
  theme(plot.title = element_text(hjust = 0.5))

```


4.  The $p$-value is an important quantity in statistics. Use a for-loop to extract the $p$-values from each test in part 2 and combine these $p$-values into a single atomic vector called `pvec_f`. Show the first 6 values.

```{r}

pvec_f <- vector(mode = "double", length = 1000)

for (i in  seq_along(pvec_f)) {
    pvec_f[i] <- tlist[[i]]$p.value
}

str(pvec_f)
head(pvec_f, n=6)

```


5.  Use the appropriate `purrr:map_*` function to extract the $p$-values from each test in part 2 and combine these $p$-values into a single atomic vector called `pvec_m`. Show the first 6 values.

```{r}

pvec_m <- map_dbl(tlist, "p.value")
str(pvec_m)
head(pvec_m, n=6)

```




6.  *Extra Credit*: $p$-values have a nice property where, if the null hypothesis is true (i.e., the mean of the observations is actually 0), then the $p$-values follow the uniform distribution.

-   Use the data from `pvec_m` to create an appropriate [QQ-plot](https://en.wikipedia.org/wiki/Q–Q_plot) and then interpret the plot with regard to whether the $p$-values exhibit a **uniform distribution.**\
-   Ensure your plot has appropriate labels for the axes and a title.
-   Include a `geom_qq_line()` that is dashed and colored red.

