---
title: "Homework 6 - Cross-Validation"
subtitle: "DATA 627 - Statistical Machine Learning, Fall 2025"
author: "Ana Paula Miranda"
date: today
number-sections: true
format:
  html:
    embed-resources: true
    toc: true
    theme: pulse
    df_print: kable
execute: 
  warning: false
---


<style>
.my-textbox {
  background-color: #f5ebeb;
  border: 1px solid #ccc;
  padding: 15px;
  border-radius: 5px;
  margin-bottom: 20px;
}
</style>

<br>


# Predicting a Grade

Do by hand. A student wants to predict their final grade for the Statistical Machine Learning course using the KNN algorithm with $K = 3$. Seven friends who took the course last year had the following mid-term test scores and grades.

<br>

* Please find the results for problem 1 on the attached pdf file.

<br>


# Cross-Validation on Simulated Data

[(Pages 222-3, chap. 5, #8)]{style="font-size: small;"}

a\. Generate a simulated data set as follows:

`set.seed(14); x <- rnorm (100); y <- x - 4*x^2 + rnorm(100);`

```{r}
suppressMessages(library(tidyverse))
library(ggplot2)
library(boot)

set.seed(14)
x <- rnorm(100)
y <- x - 4*x^2 + rnorm(100)
my_data <- data.frame(x,y)
glimpse(my_data)
```

<br>

* Write out the model used to generate the data in equation form.

The equation is:

$$ y = x -4x^2 + \epsilon$$

Where:

* $x \approx N(0, 1)$

* $\epsilon \approx N(0, 1)$, the random error.


<br>

* Plot the data and interpret the plot.


```{r}
ggplot(my_data, aes(x = x, y = y)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(se = FALSE, color = "red", size = 1.3) +
  theme_minimal()
```




<br>
<div class="my-textbox">
**Interpretation of the Plot:** 

* The scatterplot shows a parabolic (inverted U-shaped) relationship between x and y, which was expected since $x^2$ is negative, and the overall nonlinear pattern is clearly visible.

</div> 
<br>



(b) Compute the adjusted LOOCV estimates of prediction error that result from fitting each of the following four regression models:

$$
\begin{align}
    Y &= \beta_0 + \beta_1 X + \epsilon \\
    Y &= \beta_0 + \beta_1 X + \beta_2 X^2 + \epsilon \\
    Y &= \beta_0 + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 +\epsilon \\
    Y &= \beta_0 + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \beta_4 X^4 + \epsilon
\end{align}
$$

You can use the following data frame or use the `I()` function in the general linear model function formula.

```{r}
# Model 1: Linear
model1 <- glm(y ~ x, data = my_data)
cv1 <- cv.glm(my_data, model1)
cv1$delta

# Model 2: Quadratic
model2 <- glm(y ~ poly(x, 2), data = my_data)
cv2 <- cv.glm(my_data, model2)
cv2$delta

# Model 3: Cubic
model3 <- glm(y ~ poly(x, 3), data = my_data)
cv3 <- cv.glm(my_data, model3)
cv3$delta

# Model 4: 4th degree polynomial
model4 <- glm(y ~ poly(x, 4), data = my_data)
cv4 <- cv.glm(my_data, model4)
cv4$delta
```
<br>

(c) Which of these models have the smallest adjusted prediction mean squared error as estimated by LOOCV? Is this what you expected? Explain your answer.


<br>
<div class="my-textbox">
**Interpretation:** 

The quadratic model is has best fitting based on LOOCV, which matches the true nature of the data used to simulate the data.

* Model 1 (Linear) performs poorly — LOOCV error is the highest. It misses the bell shpe in the data.

* Model 2 (Quadratic) fits the best — lowest LOOCV error — agreeing with the true data structure.

* Models 3 and 4 do not significantly improve the prediction error and might slightly overfit, as seen in slightly higher prediction errors.

</div> 
<br>


d.  Use one call to `anova()` to compare all four models using the proper `test=` value and interpret the results in terms of the $p$-value.


```{r}
aout <- anova(model1, model2, model3, model4, test = "Chisq")
aout
```

<br>
<div class="my-textbox">
**Interpretation:** 

* Model 1 to Model 2: p-value < 2e-16, meaning that adding the quadratic term $x^2$ significantly improves the model.

* Model 2 to Model 3: p-value = 0.7482, meaning that Adding the cubic term $x^3$ does not significantly improved the model.

* Model 3 to Model 4: p-value = 0.7538, meaning that adding the fourth degree polinomial term $x^4$ does not significantly improved the model.

In conclusion, the quadratic model is the best fitting model. This also matches our earlier LOOCV findings.

</div> 
<br>



e.  Repeat step b but with $K=10$-Fold validation and compare the prediction mean squared error. What do you notice compared to LOOCV?

-   Use a seed of 14 where appropriate.


```{r}
# Model 1: Linear
set.seed(14)
cv1_10 <- cv.glm(my_data, model1, K = 10)
cv1_10$delta

# Model 2: Quadratic
set.seed(14)
cv2_10 <- cv.glm(my_data, model2, K = 10)
cv2_10$delta

# Model 3: Cubic
set.seed(14)
cv3_10 <- cv.glm(my_data, model3, K = 10)
cv3_10$delta

# Model 4: 4th degree polynomial
set.seed(14)
cv4_10 <- cv.glm(my_data, model4, K = 10)
cv4_10$delta
```

<br>
<div class="my-textbox">

* Model 2 is still the best fitting model here with the smallest prediction error, as expected, but it's actually larger than with LOOCV. 

* The error for model 1 is slightly smaller then that of LOOCV. 

* However, the errors for Models 2, 3 and 4 are larger than with LOOCV.

</div> 
<br>




# Predicting Defaults on Loans

[(Pages 220-221, chap. 5, #5)]{style="font-size: small;"}

<br>

Load the 'Default" data set from the {ISLR2} package.

```{r}
#| label: Default-set-up
#| message: false
#| echo: fenced
#| include: true
library(tidyverse)
library(ISLR2)
data("Default")
glimpse(Default)
```

<br>

* Use logistic regression to create a full model for predicting the probability of default based on predictors `income`, `balance`, and `student`.

```{r}
lr_fit <- glm(default ~ income + balance + student, family = binomial, data = Default)
summary(lr_fit)
```


<br>
<div class="my-textbox">
The predictors `balance` and the dummy variables for `student` status are statistically significant at 0.05 level, meaning that they are associated with the probability of default variable. However, `income` is not statistically significant in this model.
</div> 
<br>
<br>



* Use each of the following methods to estimate the *prediction error rate* of the full model and then the *prediction error rate* of a reduced model without `student`.

a.  The validation set approach. Use a 60% split for training data, a seed of 123, and a threshold of .5 where appropriate.


* With `student`:

```{r}
set.seed(123)
training_pct <- 0.6
n <-  nrow(Default)
Z <- sample(1:n, floor(n*training_pct))

dep_train <- Default[Z,]
dep_test <- Default[-Z,]

lr_fit_train <- glm(default ~ income + balance + student, 
                    family = binomial, 
                    data = dep_train)

Probs_test <- predict(object = lr_fit_train, 
                      newdata = dep_test, type = "response")

YesPredict <- 1 * (Probs_test >= 0.5)

table(YesPredict)

confusion_tab <- table(dep_test$default, YesPredict)
confusion_tab

(confusion_tab[1,1] + confusion_tab[2,2]) / sum(confusion_tab) #classification rate = accuracy
(confusion_tab[1,2] + confusion_tab[2,1]) / sum(confusion_tab) #error rate
```

<br>


* Without `student`:

```{r}
lr_fit_train_r <- glm(default ~ income + balance, 
                      family = binomial, 
                      data = dep_train)

Probs_test <- predict(object = lr_fit_train_r, 
                      newdata = dep_test, type = "response")

YesPredict <- 1 * (Probs_test >= 0.5)

table(YesPredict)

confusion_tab <- table(dep_test$default, YesPredict)
confusion_tab

(confusion_tab[1,1] + confusion_tab[2,2]) / sum(confusion_tab) #classification rate = accuracy
(confusion_tab[1,2] + confusion_tab[2,1]) / sum(confusion_tab) #error rate
```


<br>
<br>


b.  Leave-one-out cross-validation. Use a seed of 123 and a threshold of .5 for the loss function.

-   With 10,000 observations this will run for a while (mine took 5.9 minutes). Suggest including a chunk option for cache, e.g., `#| cache: true`, so it will save the results and won't run as long when you render.

```{r}
L <- function(Y, p) {
  mean(1 * (Y == 1 & p <= 0.5) | (1 * (Y == 0 & p > 0.5)),
       na.rm = TRUE)
}
```


<br>

* With `student`:

```{r}
#| cache: true
log_reg_loocv_1 <- glm(default ~ income + balance + student, 
                       family = binomial, 
                       data = Default)
set.seed(123)
cv_out_1 <- cv.glm(Default, log_reg_loocv_1, cost = L)
cv_out_1$delta
```

<br>

* Without `student`:

```{r}
#| cache: true
log_reg_loocv_2 <- glm(default ~ income + balance, 
                       family = binomial, 
                       data = Default)
set.seed(123)
cv_out_2 <- cv.glm(Default, log_reg_loocv_2, cost = L)
cv_out_2$delta
```



<br>
<br>

c.  $K$-fold cross-validation for $K = 10$ and $K = 100$. Use a seed of 123, and the same loss function with a threshold of .5 where appropriate.


* With `student`:


```{r}
#| cache: true
set.seed(123)
cv_out_10_s <- cv.glm(Default, log_reg_loocv_1, cost = L, K = 10)
cv_out_10_s$delta
```

<br>

```{r}
#| cache: true
set.seed(123)
cv_out_100_s <- cv.glm(Default, log_reg_loocv_1, cost = L, K = 100)
cv_out_100_s$delta
```


<br>


* Without `student`:


```{r}
#| cache: true
set.seed(123)
cv_out_10 <- cv.glm(Default, log_reg_loocv_2, cost = L, K = 10)
cv_out_10$delta
```

<br>

```{r}
#| cache: true
set.seed(123)
cv_out_100 <- cv.glm(Default, log_reg_loocv_2, cost = L, K = 100)
cv_out_100$delta
```


<br>


Summarize the results.

-   Compare the validation set models without and with `student` using `anova()` with the appropriate `test=`.

```{r}
aout <- anova(lr_fit_train_r, lr_fit_train, test = "Chisq")
aout
```

<br>
<div class="my-textbox">

The p-value = 0.068 is statistically significant at 0.1 level. That actually suggests that the reduced model performs better than the full model.  

</div> 
<br>



-  Interpret the results of the different methods/models and the anova test and recommend the better model.


<br>
<div class="my-textbox">
Find below the summary for the *prediction error rates* for each model tested:

* Validation set method, with `student`: 0.0265

* Validation set method, without `student`: 0.02625

* Model LOOCV, with `student`: 0.02680000

* Model LOOCV, without `student`: 0.02630000

* Model LOOCV with k = 10, with `student`: 0.0268

* Model LOOCV with k = 100, with `student`: 0.026800

* Model LOOCV with k = 10, without `student`: 0.02610

* Model LOOCV with k = 100, without `student`: 0.026300

The model with the smallest prediction error rate: 0.02610, seems to be the LOOCV with K = 10 and without `student`.

</div> 


