---
title: DATA-413/613 Homework 05 API Data
author: Ana Miranda
number-sections: true
embed-resources: true
format: html
execute:
  warning: false
  message: false
---

# Using the Smithsonian API

The Smithsonian has an [Open Access API](https://www.si.edu/openaccess/devtools) for searching its catalog with [API documentation](https://edan.si.edu/openaccess/apidocs/).

- The API returns deeply-nested lists given the complex structure of the data.
- If of interest, more detailed information can be found at [Smithsonian Institution Enterprise Data Access Network + Open Access Initiative Content Repository](https://edan.si.edu/openaccess/docs/).


**Use the {tidyverse}, {keyring}, {httr2}, and {jsonlite} packages for this question.**

```{r}

library(tidyverse)
library(keyring)
library(httr2)
library(jsonlite)
library(repurrrsive)

```


## API Key

The API requires an API key available from [data.gov](https://api.data.gov/signup/).

1.   Obtain an API key and use a {keyring} function in the console to save it in your credential manager with the name "API_KEY_DATA-GOV".

```{r}

#key_set("API_MYKEY_DATA_GOV")
#key_get("API_MYKEY_DATA_GOV")

```


## General Search Query

a.   Create a request object `search_req` for the general "search" query endpoint with URL "https://api.si.edu/openaccess/api/v1.0/search".

b. Review the [API documentation for "search"](https://edan.si.edu/openaccess/apidocs/#api-search-search) and use the `search_req` object to create a query for the phrase "artificial intelligence".
    -   You will need to add your API key to the query.
    -   Perform the query and save to an object with the name `search_resp`.

```{r}

url <- "https://api.si.edu/openaccess/api/v1.0/search"

req <- request(url) %>% 
  req_url_query(q = "artificial intelligence") %>% 
  req_url_query(rows = 10) %>% 
  req_url_query(start = 0) %>% 
  req_url_query(api_key = key_get("API_MYKEY_DATA_GOV")) %>%
  req_perform() -> search_req
str(search_req)
  
```

c. Check the status of the query response. It should be 200.

```{r}

search_req
resp_status(search_req)

```

## Convert to a Tibble with Columns of Interest

a. Convert the body of the `search_resp` object from RAW format to JSON and examine the elements of the response.
- The content of interest is in the  `response$rows` elements.
- Save the JSON output from `response$rows` to a variable called `resp_list`. It should have 10 elements.
- Convert `resp_list` to a tibble called `resp_df`. 
- It should have 10 rows with one column.

```{r}

resp_body <- resp_body_json(search_req)
glimpse(resp_body)

```

```{r}

resp_list <- resp_body$response$rows
glimpse(resp_list)

```


```{r}

resp_df <- tibble(artintel = resp_list)
glimpse(resp_df)

```


b. Use `hoist` to extract the `id` and `title` and `content` from the `resp_df$resp_list` column into columns with those names and save to a tibble called `ids_df`.

```{r}

resp_df %>% 
  hoist(
    .col = artintel,
    id ="id",
    title = "title",
    content = "content"
  ) -> ids_df
glimpse(ids_df)


```


c. Use `hoist()` to extract the element with the URL that is nested inside the `content$descriptiveNonRepeating` element and save the `ids_df`.

```{r}

ids_df %>%
  hoist(content, descriptiveNonRepeating ="descriptiveNonRepeating") %>% 
  hoist(descriptiveNonRepeating, record_link ="record_link") %>%
  select(-descriptiveNonRepeating) ->
ids_df
glimpse(ids_df)

```

d. Use `hoist()` to extract the `object_type` element that is nested inside the `content$indexStructured` element and save the `ids_df`.

```{r}

ids_df %>%
  hoist(content, indexedStructured = "indexedStructured") %>%
  hoist(indexedStructured, object_type = "object_type") %>%
  select(-indexedStructured) ->
ids_df
glimpse(ids_df)

```

e.  It is a list column. Why?

- **Answer:** Yes, it is a list and because a given title can have more than one object type.

f. Unnest the `object_type` column so the tibble has  nine variables and 10 rows and save as `ids_dfw`. 

```{r}

ids_df %>% 
  unnest_wider(object_type, names_sep = "-") -> 
ids_dfw
glimpse(ids_dfw)

```

## Smithsonian Object Types

a. Use code to show the distinct object types there are across all object columns.
- Hint: consider using a `tidyr::tidyr_tidy_select()` helper function to select just those columns and then pivot.

```{r}

ids_dfw %>% 
  select(starts_with("object_")) %>% 
  pivot_longer(
    cols = starts_with("object_"), 
    names_to = "object", 
    values_to = "type",
    names_prefix = "object_type,",
    values_drop_na = TRUE) %>% 
  group_by(type) %>% 
  count()

```


## Create a Smithsonian Search Function

a.- c. Consider if you combined the code from the above into one code chunk to create a function called `smith_search` with three arguments: `req`, `q`, and `rows` where the output is a tibble similar to what you created above with `ids_df`.

-   `req` is a request object for the search URL.
-   `q` is the query or search phrase.
-   `rows` is the number of rows in the query with a default value of 10.
-   Assume all required packages have been loaded and the API key is named "API_KEY_DATA-GOV".
-   There is no need to include documentation or error checking.

The following lines can be used in creating the function, **but they are in the wrong order and five lines are missing.** You will need to rearrange and create the missing lines.

- Note: The code lines use the natural pipe `|>` which has different rules than the magrittr pipe `%>%`.
    -   The magrittr pipe allows the use of the pronoun `.` and the natural pipe allows the use of the pronoun `_`.
    -   However, the natural pipe requires its pronoun be used with a named argument.
    -   It also requires the left side of the pipe be a completed object before using the `$` operator so it may require enclosing the left side (several steps) within parentheses.
    -   That is why you see "perhaps unusual" left or right parentheses in some of the lines below.


```{block}
    hoist(content, link = c("descriptiveNonRepeating", "record_link")) |>
    req_url_query(q = q) |>
      .col = data,
    resp_body_json(resp = _))$response$rows |>
    (req |>
       title = "title",
    tibble(data = _) |>
    hoist(
      id = "id",
    ) |>
    req_url_query(api_key = keyring::key_get("API_KEY_DATA-GOV")) |>
    unnest_wider(object_type, names_sep = "-")
      content = "content"
```


```{r}

smith_search <- function(req, q, rows) {

  (request(req) |>
    req_url_query(q = q) |>
    req_url_query(rows = rows) |> 
    req_url_query(api_key = keyring::key_get("API_MYKEY_DATA_GOV")) |> 
    req_perform() |>
     
    resp_body_json(resp = _))$response$rows |>

    tibble(data = _) |>

    hoist(
      .col = data,
      id = "id",
      title = "title",
      content = "content"
    ) |>

    hoist(content, link = c("descriptiveNonRepeating", "record_link")) |>
    hoist(content, object_type = c("indexedStructured", "object_type")) |>
    unnest_wider(object_type, names_sep = "-")
  
}    

```

d. Run your function with the arguments `search_req` and search phrase "artificial intelligence" and assign the output to `ids_test`.

```{r}

url <- "https://api.si.edu/openaccess/api/v1.0/search"

ids_test <- smith_search(url, "artificial intelligence", 10)

```


-   Use `identical()` to confirm `ids_test` is identical to `ids_dfw` for all rows of all columns except the last, without using the column number so it will work if the number of columns changes for different query phrases.

```{r}

ids_test %>% rename(artintel = data, record_link = link) -> ids_test

identical(ids_test, ids_dfw)

glimpse(ids_dfw)
glimpse(ids_test)

```
**Answer:** The data frames look identical. However, the identical() function doesn't agree. I couldn't figure out why. 

e. Run your function with the arguments `search_req`, the search phrase "artificial intelligence", and `rows = 20` and assign the output to `ids_more`.

```{r}

smith_search(url, "artificial intelligence", 20) -> ids_more
ids_more

```


g. Select the object type columns from `ids_more`. Use code to generate a tibble with a single column with only the distinct types of Smithsonian objects in all of the object type columns.
- Convert the object types to Title capitalization and sort in alphabetical order.
- Show the resulting tibble.
- It should have 12 rows.

```{r}

ids_more %>% 
  select(starts_with("object_")) %>% 
  pivot_longer(
    cols = starts_with("object_"), 
    names_to = "object", 
    values_to = "type",
    names_prefix = "object_type,",
    values_drop_na = TRUE) %>% 
  group_by(type) %>% 
  count() %>% 
  arrange(type)
  
```
**Answer:** I couldn't find 12 different categories for object type. It's looking like there are at most 10 categories available to classify an object. That could be it.


## Searching Titles (Extra Credit)

Create and show a tibble with the following three columns: 

- `words` is the output of using your function to query titles with the words "data science" with 20 rows and extracting just the titles.
- `words_single` is the output of using your function to query the titles with the phrase 'data science' inside the query with 20 rows and extracting just the titles.
- `words_double` is the output of using your function to query the titles with the phrase "data science" inside the query with 20 rows and extracting just the titles.
  - Hint: you will need to use a combination of ' and " to create a valid query with embedded phrases.
- Use {stringr} to select just the first 30 characters of each title.
- What do you observe?

```{r}

smith_search(url, "data science", 20) -> words

words %>%
  select(title) %>%
  mutate(words_single = ifelse(str_detect(title, "(?i)data|science"), title, "NA")) %>%
  mutate(words_double = ifelse(str_detect(title, "(?i)data science"), title, "NA"))
   
```

- **Answer:** It seems that requesting a query for "data science" on the Smithsonian catalog may also includes results for any titles with the word data in it, even if not related to data sciences.


# Using an API of your choice

1.  Pick a website of your choice (not discussed in class) that requires a free API key and has recent data. You may use an R package for the website or use {httr2} to interact directly with the API.\
2.  **Identify the link to instructions to obtain a key for the API.**\
3.  Use your API with {keyring} to download a data set with multiple variables. **Do not include your key in your file.**\

- **Considering for this problem FDA drug API endpoints for Adverse Drug Events:** "This is the openFDA API endpoint for adverse drug events. An adverse event is submitted to the FDA to report any undesirable experience associated with the use of a drug, including serious drug side effects, product use errors, product quality problems, and therapeutic failures."


```{r}

url_FDA <- "https://api.fda.gov/drug/event.json?"

#key_set("API_MYKEY_FDA_GOV")
#key_get("API_MYKEY_FDA_GOV")

req2 <- request(url_FDA) %>% 
  req_url_query(api_key = key_get("API_MYKEY_FDA_GOV")) %>%
  req_url_query(search = "patient.reaction.reactionmeddrapt:fatigue") %>% 
  req_url_query(limit = 10) %>% 
  req_perform() -> search_req2
str(search_req2)

```

