---
title: "Mediation Analysis"
author: "Uzoma Kanu, Ana Miranda, Jin Song"
format: html
editor: visual
---

### Mediation Analysis

**Research Question**: Does educational attainment affect residential mobility through income?

Statistical Model:

- `SCHL` → `HINCP` → `MIG`

# 1. API Querying

```{r}
# install.packages("censusapi")
library(censusapi)
library(dplyr)
```

```{r}
# API Querying

# Set your API key
# census_api_key("3cce7088b8f22e761030408d58cb4ba0cb69e9c7", install = TRUE, overwrite = TRUE)
Sys.setenv(CENSUS_API_KEY = "3cce7088b8f22e761030408d58cb4ba0cb69e9c7")

# Define the endpoint and variables
vars <- c("PWGTP", "ESR", "MAR", "MIG", "SEX", "SCH", 
          "SCHG", "SCHL", "AGEP", "HINCP", "PINCP")

# Fetch the data from ACS 2023 1-Year PUMS 
data_all <- getCensus(
  name = "acs/acs1/pums",
  vintage = 2023,
  vars = vars)
```

```{r}
# Random Sampling

# Set how many rows you want to sample (e.g., 100)
sample_size <- 50000

# Randomly sample the rows
sampled_data <- data_all[sample(nrow(data_all), sample_size), ]

# Save the data as csv
write.csv(sampled_data, "./data/sample50.csv", row.names = FALSE)
```

# 2. Load Data

```{r}
#install.packages("mediation")
#install.packages("interactions")
library(tidyverse)
library(mediation)
library(lmtest)
library(interactions)
library(ggplot2)

# Load data
data <- read.csv("./data/sample50-1.csv")
```

```{r}
# Convert variables to appropriate types
data$SCHL <- as.numeric(as.character(data$SCHL))  # Educational Attainment
data$MIG <- as.factor(data$MIG)                   # Mobility (categorical)
data$HINCP <- as.numeric(as.character(data$HINCP))# Income

# Filter out NA or irrelevant cases
data_clean <- data %>%
  filter(!is.na(SCHL), !is.na(MIG), !is.na(HINCP)) %>%
  filter(MIG %in% c("1", "2", "3"))  # Only valid mobility codes
```

# 3. Mediation Analysis

## Analysis 1. Continuous variables

```{r}
# Step 1: Mediator model
med_model <- lm(HINCP ~ SCHL, data = data_clean)

# Step 2: Outcome model with mediator
out_model <- glm(MIG ~ SCHL + HINCP, data = data_clean, family = binomial())

# Step 3: Mediation analysis
med_results <- mediate(med_model, out_model, treat = "SCHL", mediator = "HINCP", boot = TRUE)
summary(med_results)
```

### Sobel Test for continuous variables

```{r}
# adding Sobel test of significance of the indirect (mediation) effect.

# Extract coefficients and standard errors
a <- coef(summary(med_model))["SCHL", "Estimate"]
sa <- coef(summary(med_model))["SCHL", "Std. Error"]

b <- coef(summary(out_model))["HINCP", "Estimate"]
sb <- coef(summary(out_model))["HINCP", "Std. Error"]

# Sobel test statistic
sobel_z <- (a * b) / sqrt(b^2 * sa^2 + a^2 * sb^2)

# find Z critical value
critical_z <- qnorm(p = 0.05 / 2, lower.tail = FALSE)

# Two-tailed p-value
p_value <- 2 * (1 - pnorm(abs(sobel_z)))

cat("\nz-critica value =", round(critical_z, 2), "\nSobel z =", sobel_z, "\nP-value =", p_value, "\n")
```

## Analysis 2. Grouped variables

```{r}
data <- read.csv("sample50-1.csv")

# Recode Mobility into binary
data <- data %>%
  mutate(MIG_bin = case_when(
    MIG == 1 ~ 0,       # Non-mover
    MIG %in% c(2, 3) ~ 1))

data$MIG_bin <- as.numeric(data$MIG_bin)
```

```{r}
# Recode Educational Attainment (Groups of Low, Medium, High)
data <- data %>%
  mutate(SCHL_bin = case_when(
    SCHL >= 1 & SCHL <= 15 ~ "Low",
    SCHL >= 16 & SCHL <= 20 ~ "Medium",
    SCHL >= 21 & SCHL <= 24 ~ "High"))

# Recode Income into quantiles
quantiles <- quantile(data$HINCP, probs = c(0.33, 0.66), na.rm = TRUE)
data <- data %>%
  mutate(HINCP_bin = case_when(
    HINCP <= quantiles[1] ~ "Low",
    HINCP > quantiles[1] & HINCP <= quantiles[2] ~ "Medium",
    HINCP > quantiles[2] ~ "High"))

# Remove missing data
data <- data %>% filter(!is.na(SCHL_bin), !is.na(HINCP_bin), !is.na(MIG_bin))

# Convert to factors
data$SCHL_bin <- factor(data$SCHL_bin, levels = c("Low", "Medium", "High"))
data$HINCP_bin <- factor(data$HINCP_bin, levels = c("Low", "Medium", "High"))

# Convert HINCP_bin to numeric for mediation
data$SCHL_num <- as.numeric(data$SCHL_bin)
data$HINCP_num <- as.numeric(data$HINCP_bin)
```

```{r}
# Mediator model (Education → Income)
med_model <- lm(HINCP_num ~ SCHL_num, data = data)

# Outcome model (Education + Income → Mobility)
out_model <- glm(MIG_bin ~ SCHL_num + HINCP_num, family = binomial(), data = data)

# Mediation analysis
set.seed(123)
med_out <- mediate(med_model, out_model,
                   treat = "SCHL_num", mediator = "HINCP_num", sims = 1000)
summary(med_out)
```

```{r}
plot(med_out) # Plot mediation effects
```

### Sobel Test for Grouped variables

```{r}
# adding Sobel test of significance of the indirect (mediation) effect.

# Extract coefficients and standard errors
a <- coef(summary(med_model))["SCHL_num", "Estimate"]
sa <- coef(summary(med_model))["SCHL_num", "Std. Error"]

b <- coef(summary(out_model))["HINCP_num", "Estimate"]
sb <- coef(summary(out_model))["HINCP_num", "Std. Error"]

# Sobel test statistic
sobel_z <- (a * b) / sqrt(b^2 * sa^2 + a^2 * sb^2)

# find Z critical value
critical_z <- qnorm(p = 0.05 / 2, lower.tail = FALSE)

# Two-tailed p-value
p_value <- 2 * (1 - pnorm(abs(sobel_z)))

cat("\nz-critica value =", round(critical_z, 2), "\nSobel z =", sobel_z, "\nP-value =", p_value, "\n")
```

- Interpretation of the Sobel significance test:

  - A large absolute Sobel z-value, usually larger than the z-critical value, indicates mediation effect in significant at 0.05 level.
  - In the example above, z = -20.40 and it is significant because it's larger than z-critical = 1.96.
  - With a p-value = 0, we can reject the hull hypothesis that the indirect mediation effect is zero, meaning that there is a significant mediation effect.
  - In conclusion: There is sufficient evidence of a significant indirect effect of schooling (SCHL) on migration status (MIG) through household income (HINCP), suggesting that education influences migration because it affects income.

# 4. Moderation Analysis

```{r}


# Convert SEX to binary (0 = Male, 1 = Female)
data_clean <- data_clean %>%
  mutate(SEX = ifelse(SEX == 1, 0, 1)) 

```

```{r}


# Updated mediator model with interaction
med_model_mod <- lm(HINCP ~ SCHL * SEX, data = data_clean)

# Outcome model remains unchanged
out_model <- glm(MIG ~ SCHL + HINCP, 
                data = data_clean, family = binomial())

# Mediation analysis with moderation
set.seed(123)
med_out_mod <- mediate(med_model_mod, out_model,
                      treat = "SCHL", mediator = "HINCP",
                      boot = TRUE, sims = 1000)
summary(med_out_mod)


```

```{r}


# Mediator model remains unchanged
med_model <- lm(HINCP ~ SCHL, data = data_clean)

# Updated outcome model with interaction
out_model_mod <- glm(MIG ~ SCHL + HINCP * SEX,
                    data = data_clean, family = binomial())

# Mediation analysis with moderation
set.seed(123)
med_out_mod <- mediate(med_model, out_model_mod,
                      treat = "SCHL", mediator = "HINCP",
                      boot = TRUE, sims = 1000)
summary(med_out_mod)

```

```{r}

library(interactions)

# For a-path moderation
interact_plot(med_model_mod, pred = SCHL, modx = SEX,
             interval = TRUE, int.type = "confidence")

# For b-path moderation
interact_plot(out_model_mod, pred = HINCP, modx = SEX,
             interval = TRUE, int.type = "confidence")



```
