---
title: "DATA 615 - Regression: Homework - 1"
author: "Ana Miranda"
date: " `r Sys.Date()` "
output:
  # word_document: 
  #   toc: yes
  #   toc_depth: 2
  #   number_sections: no
  pdf_document: 
    toc: yes
    toc_depth: 2
    number_sections: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(ggthemes)
library(patchwork)
```


## Q.1. (5 pts)

Yes, "I have read Lab 1. I have installed relevant software, and am ready to use software for homework, labs, and data analysis."

## Q.2, Problem 1.20  (10 pts)

```{r}
# Load data
cm <- read.table("./data1/CH01PR20.txt", header = F)
colnames(cm) <- c("time", "copiers")
glimpse(cm)

summary(cm)

```

Observation: For each call, X is the number of copiers serviced and Y is the total number of minutes spent by the service person.


### Graphical summary (optional, not part of HW)

```{r}

ggplot(cm, mapping = aes(x = time)) +
  geom_histogram(fill = "gray", color = "black") +   theme_bw() -> g1

ggplot(cm, mapping = aes(x = copiers)) +
  geom_histogram(fill = "gray", color = "black") +   theme_bw() -> g2

(g1 + g2)

```


- Findings:

    - Variable `time` seems to be somewhat uniformally distributed and with no visible outliers.
    - Variable `copiers` seems to be a little skewed to the right and with no visible outliers.
      

```{r}

ggplot(cm, mapping = aes(x = copiers, y = time)) +
  geom_point() +
  theme_bw()

```

- Findings:

    - The variables `time` and `copiers` present a positive and strong linear corretation. I.e, as `copiers` increase, `time` also increases.
    
    

### a. Estimated regression function (You do NOT have to copy the questions. I often include a brief description for my convenience.)


```{r}

cm_SLR <- lm(data = cm, time ~ copiers)
summary(cm_SLR)

```

The estimated regression line (function) is:

$$\widehat{time} = -0.5802 + 15.0352(copiers)$$

### b. Plot data with regression line

```{r}

ggplot(cm, mapping = aes(x = copiers, y = time)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, colour = "red") +
  theme_bw()

```

The above plot shows indeed that the variables `copiers` and `time` hold a strong linear relationship. In fact, based on the linear model adjusted $R-squared = 0.9565$, 95% of the total variation in `time`, can be explained by the simple linear regression with `copiers` as the predictor.


### c. Interpret the intercept

- The estimated intercept is $\beta_0 = -0.5802$. This means the predicted value of `time` when the value of `copiers` is equal to zero.

- The estimated slope is positive. This means that for every unit increase in `copiers`, the `time` is expected to increase 15.0352.



### d. Predict y at x = 5.

$$\widehat{time} = -0.5802 + 15.0352(5) = 74.60$$

## Q.3. Problem 1.24 (5 pts)

### a. Residuals and Sum of the squared residuals

```{r}
cm_SLR$residuals

anova(cm_SLR)
```

As provided by the ANOVA output above, the sum squared of the residuals is 3.416 with 43 degrees of freedom.


### b. Estimate $\sigma^2$ and $\sigma$.

The estimated variance of the error is $MSE =  3.416/43 = 79.45$. Therefore, the estimated standard deviation of the model is $\sqrt{MSE} = \sqrt{79.45} =  8.9135$.


```{r}

install.packages("ggfortify")
library(ggfortify)

autoplot(cm_SLR)

library(broom)
df <- augment(cm_SLR)
df
ggplot(df, aes(x = .fitted, y = .resid)) + geom_point()

ls(cm_SLR)

ggplot(cm_SLR, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0)

ggplot(cm_SLR, aes(sample = .resid)) +
  geom_qq() +
  geom_qq_line() +
  #geom_abline(color = "red") +
  geom_abline(color = "red", linetype = "dashed") +
    ggtitle("Normal Q-Q Plot of Residuals") +
  theme_bw()

```

