---
title: "Final Project Analysis"
subtitle: "DATA 627 - Statistical Machine Learning, Fall 2025"
author: "Ana Paula Miranda"
author2: "Andrew McPherson"
date: today
number-sections: true
format:
  html:
    embed-resources: true
    toc: true
    theme: pulse
    df_print: kable
execute: 
  warning: false
---

# Data Preparation

## Libraries

```{r libraries}
#| message: false
#| echo: fenced
#| include: true
library(tidyverse)
library(forcats)
library(lubridate)
library(randomForestSRC)
library(caret)
library(class)
library(boot)
library(ggplot2)
library(tree)
options(scipen = 999)
```

## Set Seed

```{r}
myseed = 1234
```


## Load Data

```{r load}
df = read_csv("./data/apprentices_2018_2019.csv", col_types = cols())
app = df

glimpse(app)
```

## Transform Data

```{r transformations}
app = app %>% 
  select(APPR_ID, 
         MAJOR_GROUP_NAME, 
         ONET_SOC_CD, 
         NAICS_CD, 
         APPR_STATUS_CD, 
         START_DT, 
         STARTING_WAGE, 
         EXIT_WAGE_DT, 
         EXIT_WAGE, 
         EXPECT_COMPLETE_DT,
         COMPLETION_EXT_DT, 
         AGE_AT_START, 
         SEX_CD, 
         ETHNICITY_CD, 
         EDUCATION_CD, 
         RACE_CD, 
         ASIAN_IND, 
         AFRICAAMER_IND, 
         NATHAWAI_IND, 
         NATINDIAN_IND, 
         W_IND, 
         OCCUP_TYPE
         ) %>% 
  mutate(
    
  STARTING_WAGE = ifelse(STARTING_WAGE < 5.00, NA, STARTING_WAGE),
  STARTING_WAGE = ifelse(STARTING_WAGE > 80, NA, STARTING_WAGE),

  EXIT_WAGE = ifelse(EXIT_WAGE < 5.00, NA, EXIT_WAGE),
  EXIT_WAGE = ifelse(EXIT_WAGE > 80, NA, EXIT_WAGE),


  COMPLETION_EXT_DT = ifelse(COMPLETION_EXT_DT < START_DT, NA, COMPLETION_EXT_DT),
  COMPLETION_EXT_DT = ifelse(COMPLETION_EXT_DT > as.Date("2025-06-30"), NA, COMPLETION_EXT_DT),
  COMPLETION_EXT_DT = as.Date(COMPLETION_EXT_DT),

# The Office of Apprenticeship defines successful completion as an apprentice completing their apprenticeship within 1 year of the expected completion date. Treating completion in this way allows OA to define a status for the apprentices who left their apprenticeship but did not have their status updated by their sponsor.   
    
  COMPLETION_DEADLINE = case_when(
    EXPECT_COMPLETE_DT < START_DT ~ as.Date(NA),
    TRUE ~ as.Date(EXPECT_COMPLETE_DT + years(1))
),

# From Data Dictionary Addendum. APPRENTICE_STATUS_CODE: Status of the Apprentice, this determines if an apprentice, is Active, Cancelled, New or Completed. CA = Cancelled, CO = Completed, RE = Registered, RI = Reinstated, SU = Suspended, TR = Transferred, RS = Restore Pending, UP = Update Pending/Registered, IC = Interim Completion Pending, RS=Restore Pending. 
# Given data quality issues, we must be careful. We will treat an apprentice with CO as completed, and an apprentice with CA as cancelled. However, it is possible that a sponsor has not updated this field. Therefore we will also make a determination based on the value of completion date, since the presence of a completion date should be definitive as a completion event. 
  
  COMPLETION = case_when(
    APPR_STATUS_CD == "CO" ~ "Y",
    APPR_STATUS_CD == "CA" ~ "N",
    is.na(COMPLETION_DEADLINE) ~ NA,
    COMPLETION_DEADLINE > as.Date("2025-06-30") ~ NA, 
    is.na(COMPLETION_EXT_DT) ~ "N",
    COMPLETION_EXT_DT > COMPLETION_DEADLINE ~ "N",
    COMPLETION_EXT_DT <= COMPLETION_DEADLINE ~ "Y"
  ),
  COMPLETION = as.factor(COMPLETION),
  
  MAJOR_GROUP_NAME = as.factor(MAJOR_GROUP_NAME),
  APPR_STATUS_CD = as.factor(APPR_STATUS_CD), 
  SEX_CD = as.factor(SEX_CD),
  ETHNICITY_CD = as.factor(ETHNICITY_CD),
  EDUCATION_CD = as.factor(EDUCATION_CD),
  RACE_CD = as.factor(RACE_CD),
  ASIAN_IND = as.factor(ASIAN_IND),
  AFRICAAMER_IND = as.factor(AFRICAAMER_IND),
  NATHAWAI_IND = as.factor(NATHAWAI_IND),
  NATINDIAN_IND = as.factor(NATINDIAN_IND),
  W_IND = as.factor(W_IND),
  OCCUP_TYPE = as.factor(OCCUP_TYPE),
  
  AGE_AT_START = ifelse((AGE_AT_START < 14 | AGE_AT_START > 70), NA, AGE_AT_START),

# 
#   EDUCATION_CD_Rec = ifelse(is.na(EDUCATION_CD), "5", EDUCATION_CD),
#   EDUCATION_CD_Rec = as.factor(EDUCATION_CD_Rec),
# 
#   SEX_CD_Rec = ifelse(is.na(SEX_CD), "3", SEX_CD),
#   SEX_CD_Rec = as.factor(SEX_CD_Rec),

  # ETHNICITY_CD_Rec = ifelse(is.na(ETHNICITY_CD), "3", ETHNICITY_CD),
  # ETHNICITY_CD_Rec = as.factor(ETHNICITY_CD_Rec),
  # 
  # RACE_CD_Rec = ifelse(is.na(RACE_CD), "6", RACE_CD),
  # RACE_CD_Rec = as.factor(RACE_CD_Rec),
     
  ) %>% 

  relocate(COMPLETION_DEADLINE, .after = COMPLETION_EXT_DT)  %>% 
  relocate(COMPLETION, .after = COMPLETION_DEADLINE)

  
glimpse(app)

app = app %>% 
  mutate(
    RACE_CD = fct_recode(RACE_CD,
                          "White" = "1",
                          "Pacific Islander" = "2",
                          "Black" = "3",
                          "Asian" = "4",
                          "Native American" = "5",
                          "Not Provided" = "6",
                          "Multi-Race" = "7"
    ),
    EDUCATION_CD = fct_recode(EDUCATION_CD,
                              "Not High School graduate"  = "1",
                              "9th to 12th grade, no diploma"  = "2",
                              "GED"  = "3",
                              "High School graduate (including equivalency)"  = "4",
                              "Participant Did Not Self-Identify"  = "5",
                              "Some College or Associate's degree"  = "6",
                              "Bachelor's degree"  = "7",
                              "Masters degree"  = "8",
                              "Doctorate or Prof. degree"  = "9"
    ),
    ETHNICITY_CD = fct_recode(ETHNICITY_CD,
                              "Hispanic or Latino"  = "1",
                              "Non-Hispanic or Latino"  = "2",
                              # "Did Not Self-Identify"  = "3",
                              "Did Not Self-Identify"  = "9"
    ),
    SEX_CD = fct_recode(SEX_CD,
                              "Male"  = "1",
                              "Female"  = "2",
                              # "Did Not Self-Identify"  = "3",
                              "Did Not Self-Identify"  = "9"
    )
  )  

glimpse(app)
```

## Re-coding Race Indicators and `MAJOR_GROUP_NAME`

```{r}
app_final_reg = app %>% 
  mutate(
    ASIAN_dummy = case_when(
       ASIAN_IND == "1" ~ 1,
       ASIAN_IND == "0" ~ 0,
       ASIAN_IND == "2" ~ 0, 
       ASIAN_IND == "9" ~ 9,
       is.na(ASIAN_IND) ~ 9,
       TRUE ~ NA),
    
    AFRICAAMER_dummy = case_when(
       AFRICAAMER_IND == "1" ~ 1,
       AFRICAAMER_IND == "0" ~ 0,
       AFRICAAMER_IND == "2" ~ 0, 
       AFRICAAMER_IND == "9" ~ 9,
       is.na(AFRICAAMER_IND) ~ 9,
       TRUE ~ NA),

    NATHAWAI_dummy = case_when(
       NATHAWAI_IND == "1" ~ 1,
       NATHAWAI_IND == "0" ~ 0,
       NATHAWAI_IND == "2" ~ 0, 
       NATHAWAI_IND == "9" ~ 9,
       is.na(NATHAWAI_IND) ~ 9,
       TRUE ~ NA),

    NATINDIAN_dummy = case_when(
       NATINDIAN_IND == "1" ~ 1,
       NATINDIAN_IND == "0" ~ 0,
       NATINDIAN_IND == "2" ~ 0, 
       NATINDIAN_IND == "9" ~ 9,
       is.na(NATINDIAN_IND) ~ 9,
       TRUE ~ NA),

    WHITE_dummy = case_when(
       W_IND == "1" ~ 1,
       W_IND == "0" ~ 0,
       W_IND == "2" ~ 0, 
       W_IND == "9" ~ 9,
       is.na(W_IND) ~ 9,
       TRUE ~ NA),
    
    ASIAN_dummy = as.factor(ASIAN_dummy),
    AFRICAAMER_dummy = as.factor(AFRICAAMER_dummy),
    NATHAWAI_dummy = as.factor(NATHAWAI_dummy),
    NATINDIAN_dummy = as.factor(NATINDIAN_dummy),
    WHITE_dummy = as.factor(WHITE_dummy),
    
    ASIAN_dummy = fct_recode(ASIAN_dummy, 
                             "Asian" = "1",
                             "Non-Asian" = "0",
                             "Did Not Self-Identify" = "9"
    ),
    AFRICAAMER_dummy = fct_recode(AFRICAAMER_dummy, 
                                  "African American" = "1",
                                  "Non-African American" = "0",
                                  "Did Not Self-Identify" = "9"
    ),
    NATHAWAI_dummy = fct_recode(NATHAWAI_dummy, 
                                "Native Hawaiian or Pacific Islander" = "1",
                                "Non-Native Hawaiian or Pacific Islander" = "0",
                                "Did Not Self-Identify" = "9"
    ),
    NATINDIAN_dummy = fct_recode(NATINDIAN_dummy, 
                                 "American Indian or Alaska Native" = "1",
                                 "Non-American Indian or Alaska Native" = "0",
                                 "Did Not Self-Identify" = "9"
    ),
    WHITE_dummy = fct_recode(WHITE_dummy, 
                             "White" = "1",
                             "Non-White" = "0",
                             "Did Not Self-Identify" = "9"
    ),
    MAJOR_GROUP_NAME = fct_recode(MAJOR_GROUP_NAME,
                                  "STEM and Technical Professions" = "Architecture and Engineering Occupations",
                                  "STEM and Technical Professions" = "Computer and Mathematical Occupations",
                                  "STEM and Technical Professions" = "Life, Physical, and Social Science Occupations",
                                  
                                  "Trades and Manual Labor" = "Construction and Extraction Occupations",
                                  "Trades and Manual Labor" = "Installation, Maintenance, and Repair Occupations",
                                  "Trades and Manual Labor" = "Production Occupations",
                                  "Trades and Manual Labor" = "Transportation and Material Moving Occupations",
                                  "Trades and Manual Labor" = "Farming, Fishing, and Forestry Occupations",
                                  
                                  "Healthcare and Social Services" = "Healthcare Practitioners and Technical Occupations",
                                  "Healthcare and Social Services" = "Healthcare Support Occupations",
                                  "Healthcare and Social Services" = "Community and Social Service Occupations",
                                  "Healthcare and Social Services" = "Personal Care and Service Occupations",
                                  
                                  "Education and Public Services" = "Educational Instruction and Library Occupations",
                                  "Education and Public Services" = "Protective Service Occupations",
                                  "Education and Public Services" = "Military Specific Occupations",
                                  
                                  "Business, Management and Office Support" = "Management Occupations",
                                  "Business, Management and Office Support" = "Business and Financial Operations Occupations",
                                  "Business, Management and Office Support" = "Office and Administrative Support Occupations",
                                  
                                  "Arts, Hospitality and Sales" = "Arts, Design, Entertainment, Sports, and Media Occupations",
                                  "Arts, Hospitality and Sales" = "Food Preparation and Serving Related Occupations",
                                  "Arts, Hospitality and Sales" = "Sales and Related Occupations",
                                  "Arts, Hospitality and Sales" = "Building and Grounds Cleaning and Maintenance Occupations"
    )
      )


# Suggestive Re-coding of MAJOR_GROUP_NAME into 6 broader categories below:
# 1. STEM & Technical Professions: grouped occupations centered around science, technology, engineering, and mathematics.
# 2. Trades & Manual Labor: grouped occupations around hands-on, skilled labor and physical work.
# 3. Healthcare & Social Services: grouped roles focused on health, wellness, and social support.
# 4. Education & Public Services: grouped jobs related to education, public safety, and government services.
# 5. Business, Management & Office Support: grouped administrative, financial, and managerial roles.
# 6. Arts, Hospitality & Sales: grouped creative, customer-facing, and service-oriented roles.


app_final_reg = app_final_reg %>% 
  select(
    MAJOR_GROUP_NAME,
    STARTING_WAGE   ,
    EXIT_WAGE       ,  
    COMPLETION      ,
    AGE_AT_START    ,
    OCCUP_TYPE      ,
    EDUCATION_CD,
    SEX_CD      ,
    ETHNICITY_CD,
    RACE_CD,
    ASIAN_dummy     ,
    AFRICAAMER_dummy,
    NATHAWAI_dummy  ,
    NATINDIAN_dummy ,
    WHITE_dummy     
      )

app_final_reg_no_na <- na.omit(app_final_reg)

table(app_final_reg$MAJOR_GROUP_NAME, useNA = "always")
```

# Data Exploration 

## Qualitative Predictors

### Examine Qualitative Predictors

```{r}
glimpse(app_final_reg %>% 
  select(where(is.factor) | where(is.character)))
```

### Determine Number of Levels for Predictors

```{r}
app_final_reg %>%
  select(where(is.factor), where(is.character)) %>%
  sapply(nlevels)
```
### Qualitative Predictor Analysis

#### MAJOR_GROUP_NAME

##### Distribution

```{r}
app_final_reg %>%
  count(MAJOR_GROUP_NAME) %>%
  ggplot(aes(x = reorder(MAJOR_GROUP_NAME, n), y = n)) +
  geom_col(fill = "maroon") +
  coord_flip() +
  labs(x = "Occupational Groups", y = "Count", title = "Distribution of Major Occupational Groups")
```

##### Relationship with Target Variables

```{r}
table(app_final_reg$MAJOR_GROUP_NAME, app_final_reg$COMPLETION)
prop.table(table(app_final_reg$MAJOR_GROUP_NAME, app_final_reg$COMPLETION), margin = 1)
```

```{r}
library(ggplot2)
ggplot(app_final_reg, aes(x = MAJOR_GROUP_NAME, fill = COMPLETION)) +
  geom_bar(position = "fill") +
  labs(y = "Proportion Completed") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
##### Association Tests

```{r}
# Cramer's V Test
vcd::assocstats(table(app_final_reg$MAJOR_GROUP_NAME, app_final_reg$COMPLETION))

# Chi-square Test
chisq.test(table(app_final_reg$MAJOR_GROUP_NAME, app_final_reg$COMPLETION))
```

:::{.callout-note}
## Occupational Group Distribution
The data suggests that Occupational group has a relationship with Completion status, with significant differences between factor levels. While the Cramer's V score is fairly low, the chi-square test has a very low p-value, indicating that the null hypothesis (that the two variables are independent) is unlikely. The large number of NA values present difficulties.
:::

#### COMPLETION

```{r}
app_final_reg %>%
  count(COMPLETION) %>%
  ggplot(aes(x = reorder(COMPLETION, n), y = n)) +
  geom_col(fill = "maroon") +
  coord_flip() +
  labs(x = "Completion Status", y = "Count", title = "Distribution of Completion")
```

:::{.callout-note}
## Completion Distribution
The dataset is broadly split between apprentices who did and did not complete. There is a relatively low number of NAs, and many of these are meaningful (i.e. apprentices who are not expected to have completed yet.)
:::

#### SEX_CD

```{r}
app_final_reg %>%
  count(SEX_CD) %>%
  ggplot(aes(x = reorder(SEX_CD, n), y = n)) +
  geom_col(fill = "maroon") +
  coord_flip() +
  labs(x = "Sex Code", y = "Count", title = "Distribution by Sex")
```

:::{.callout-note}
## Sex Distribution
The dataset is strongly skewed towards male participants. This may impact the results in some methods, particularly tree-based methods. These methods can disregard a variable if it doesn't have a strong overall influence on the dependent variable prediction. This can happen when the data is skewed, even when there is a significant difference between the classification outcomes between the variables.  
:::

#### ETHNICITY_CD

```{r}
app_final_reg %>%
  count(ETHNICITY_CD) %>%
  ggplot(aes(x = reorder(ETHNICITY_CD, n), y = n)) +
  geom_col(fill = "maroon") +
  coord_flip() +
  labs(x = "Ethnicity Code", y = "Count", title = "Distribution by Ethnicity")
```

#### EDUCATION_CD

```{r}
app_final_reg %>%
  count(EDUCATION_CD) %>%
  ggplot(aes(x = reorder(EDUCATION_CD, n), y = n)) +
  geom_col(fill = "maroon") +
  coord_flip() +
  labs(x = "Education Code", y = "Count", title = "Distribution by Education")
```

:::{.callout-note}
## Education Distribution
The data here is unbalanced, which should be considered when Education is included as a predictor. 
:::

#### OCCUP_TYPE

```{r}
app_final_reg %>%
  count(OCCUP_TYPE) %>%
  ggplot(aes(x = reorder(OCCUP_TYPE, n), y = n)) +
  geom_col(fill = "maroon") +
  coord_flip() +
  labs(x = "Training Model Code", y = "Count", title = "Distribution by Training Model")
```

:::{.callout-note}
## Training Model Distribution
The Training Model code is dominated by NAs. Furthermore, Time-Based training models vastly outweigh Time-based and Hybrid. It may not be advisable to use this code as a predictor. However, some methods, such as random forest, manage NA values well, so we can explore whether to include it. 
:::


## Quantitative Predictors

### Examine Quantitative Predictors

```{r}
glimpse(app_final_reg %>% 
  select(where(is.numeric)))
```

#### Correlation Tests

```{r}
app_final_reg_no_na[, purrr::map_lgl(app_final_reg, is.numeric)] |> 
  cor() |> round(2)

corrr::correlate(app_final_reg[, purrr::map_lgl(app_final_reg, is.numeric)], quiet = TRUE) |> ggplot2::autoplot()
```

* **The Correlations**

  * EXIT_WAGE vs STARTING_WAGE: Corr = 0.644*, indicating a strong positive correlation. Higher starting wages tend to lead to higher exit wages.

  * EXIT_WAGE vs AGE_AT_START: Corr = -0.064*, suggesting a very weak negative correlation. Older age slightly associates with lower exit wages.
  
  * STARTING_WAGE vs AGE_AT_START: Corr = 0.032*, suggesting a very weak positive correlation. Essentially no meaningful correlation.



#### Distributions

##### STARTING_WAGE

```{r}
#| message: false
#| warning: false
library(scales)
ggplot(app_final_reg, 
       aes(x = STARTING_WAGE, y= after_stat(count/sum(count)))) +
  geom_histogram(fill = "cornflowerblue", 
                 color = "white") + 
  labs(title="Apprentices Wage at Start", 
       y = "Percent",
       x = "Starting Wage") +
  scale_y_continuous(labels = percent)
```

##### EXIT_WAGE

```{r}
#| message: false
#| warning: false
ggplot(app_final_reg, aes(x = EXIT_WAGE, y= after_stat(count/sum(count)))) +
  geom_histogram(fill = "cornflowerblue", 
                 color = "white") + 
  labs(title="Apprentices Wage at Exit", 
       y = "Percent",
       x = "Exit Wage")  +
  scale_y_continuous(labels = percent)
```


##### EXIT_WAGE by MAJOR_GROUP_NAME

```{r}
#| message: false
#| warning: false
library(ggthemes)
ggplot(app_final_reg, aes(x = EXIT_WAGE)) +
  geom_histogram() +
  facet_wrap(~MAJOR_GROUP_NAME, ncol = 3) +
  labs(title = " ",
       x = " ") +
  theme_fivethirtyeight()
```


##### AGE_AT_START

```{r}
#| message: false
#| warning: false
library(scales)
ggplot(app_final_reg, 
       aes(x = AGE_AT_START)) +
  geom_histogram(fill = "cornflowerblue", 
                 color = "white") + 
  labs(title="Apprentices Staring Age", 
       y = "Count",
       x = "Age") +
  scale_y_continuous()
```

* **The Distributions**

  * STARTING_WAGE and EXIT_WAGE are similarly distributed. The distribution is right-skewed (positively skewed), meaning most apprentices start with relatively low wages, and only a few start at very high wages. The majority of starting wages fall between 10 and 30 dollars per hour.

  * AGE_AT_START: Right-skewed distribution which then declines gradually, meaning most apprentices start young, and fewer start at older ages.
The highest concentration of apprentices is between 20 and 35 years old. It shows that Apprenticeships primarily attract young adults, especially those in their early 20s. 



# Analysis

## Regression

The objective of this study is to determine which features are the most relevant in predicting post-apprenticeship wage outcomes to potentially guide program designs and apprentice’s expectations regarding economic advancement. The purpose of this analysis is prediction, rather than inference.

My research question is: Is it possible to predict post-apprenticeship wages using program attributes and apprentice demographics, including occupational group, training model, initial wage, age, gender, educational level, and race?

### Data Set

```{r}
clean_data_wage_reg = drop_na(app_final_reg) %>% 
  select(-ASIAN_dummy, -AFRICAAMER_dummy, -NATHAWAI_dummy, -NATINDIAN_dummy, -WHITE_dummy, -COMPLETION)
glimpse(clean_data_wage_reg)
```


### Validation set with 70%/30% split

```{r}
n <- nrow(clean_data_wage_reg)
set.seed(myseed)
training_pct <- 0.7
w_Z <- sample(1:n, floor(n*training_pct))
reg_train <- clean_data_wage_reg[w_Z,]
reg_test <- clean_data_wage_reg[-w_Z,]
```


### Multiple Linear Regression

```{r}
mlr_wage <- lm(EXIT_WAGE ~ STARTING_WAGE + AGE_AT_START + SEX_CD +
               ETHNICITY_CD + EDUCATION_CD + RACE_CD + OCCUP_TYPE + MAJOR_GROUP_NAME,
               reg_train)
summary(mlr_wage)

mlr_wage_predict = predict(mlr_wage, reg_test)
mean((reg_test$EXIT_WAGE - mlr_wage_predict)^2)
sqrt(mean((reg_test$EXIT_WAGE - mlr_wage_predict)^2))
```

#### Model Fit {-}

Adjusted $R^2$ = 0.4449, so this model explains about 44.5% of the variance in exit wage, which is quite reasonable for wage data. For a dataset this large, n = 50,099, even modest effects become highly statistically significant - so effect size matters more than p-values.

The residual standard error = 9.2, This means the prediction error for EXIT_WAGE is about $\pm 9.2$ wage units (dollars/hour).

Test RMSE = 9.28, meaning that our out-of-sample mean squared error is very close to the training RSE, which indicates:

* Very little overfitting.

* The model generalizes well to new data.

* The linear model is performing as expected.


##### Key Predictors and Interpretation {-}

* Starting wage: STARTING_WAGE ($\beta = 1.22$), is a strong predictor. For every one unit increase in starting wage, the exit wage increases by 1.22 dollars/hour, which can be considered a very large and meaningful effect. This alone drives much of the model’s predictive power, meaning that people tend to leave their job with a wage that is strongly associated to the wage they started with.

* Age: AGE_AT_START ($\beta \approx -0.09$), has a small, but significant negative effect. Meaning that for each additional year of age decreases exit wage by approximately 9 cents. This is not a large practical impact, but statistically significant.

* Sex, SEX_CD:

  * Female workers ($\beta \approx -0.913$) have about $0.913 lower exit wage, all else equal, compared to males.


* Ethnicity, ETHNICITY_CD:

  * Non-Hispanic workers ($\beta \approx -0.63$) have about $0.63 lower exit wage, all else equal, compared to Hispanics.
  
  * The workers who did not self-identify an ethnicity ($\beta \approx -1.28$) have about $1.28 lower exit wage, all else equal, compared to Hispanics.


* Education Effects, EDUCATION_CD, compared to the reference level - "Not High School graduate":

  * GED: + $2.62

  * High School graduate (including equivalency): + $1.70

  * Some College or Associate's degree: + $1.74
  
  * Bachelor's degree: + $1.62 (not significant)

  * Masters: + $2.57 (not significant)

Higher education generally increases exit wages, but the significance and magnitude of such effect vary. It's probably due to job type differences or sample size issues within the categories.


* Race, RACE_DC, compared to the reference level - "White":

  * Black workers: − $2.46, with a strong negative difference in exit wage.

  * Asian workers: + $1.61, with a strong positive difference in exit wage.

  * Native American: + $1.32, with positive significant in exit wage.

  * The other categories are not statistically meaningful.


* Occupational type, OCCUP_TYPE, compared to the reference level - "Competency Based":

  * Hybrid jobs: + $1.00

  * Time-based jobs: + $1.09


* Major occupational group, MAJOR_GROUP_NAME, compared to reference level - "STEM and Technical Professions". The following differences meaningfully impact exit wages. 

  * Trades and Manual Labor: + $1.96, this is the strongest positive effect.

  * Education and public services: − $2.97, this has a very strong negative effect in exit wage.

  * Healthcare and Social Services: - $1.43.
  

##### Conclusion: {-}

The multiple linear regression model explains about 45% of the variation in exit wages, with a residual prediction error of about +-9 dollars/hour and a test-set RMSE of 9.28, indicating strong generalization (predictive power) and minimal overfitting.

* Starting wage is by far the strongest predictor, increasing exit wage by about $1.20 dollar/hour of starting wage. Age has a small negative effect.

* Several demographic factors show statistically significant differences:

* Females have lower exit wages (− $0.913),

* Black workers earn less (− $2.46),

* Asian workers earn more (+ $1.61),

* Education levels also affect exit wages, with higher education generally increasing wages.

* Occupational and major group categories also matter, with "Trades/Manual Labor roles" having the highest positive association and "Education/Public Service" jobs the most negative.

Overall, the model captures meaningful patterns but leaves room for additional nonlinear or interaction effects to improve predictive accuracy. Next, I'll ML methods to capture more complex patterns in the data.


```{r}
library(broom)
library(kableExtra)
tidy_model <- tidy(mlr_wage)
kable(tidy_model, format = "html") %>%
  kable_styling("striped", full_width = F)

```



#### Check for multicollinearity - Variance Inflation Factor (VIF)

```{r}
library(car)
vif(mlr_wage)
``` 

The VIF results show no evidence of multicollinearty between the predictor since all the adjusted GVIF values fall between 1.02 and 1.19 - far below any concerning threshold. This means that the coefficient estimates for this model are stable, the standard errors are reliable and that the interpretation of individual predictors is meaningful.

### Model Selection

#### Leave-One-Out Cross-Validation (LOOCV).

I'm not going to run this analysis for this data set at this point because we have a large number of observations and the computer is able to handle it well. Further discussion may be necessary while talking into consideration other methods.

```{r}
# #| cache: true
# library(boot)
# loocv_wage_full = cv.glm(clean_data_wage_reg, mlr_wage_full)
# loocv_wage_full$delta
```


#### k-Fold Cross-Validation

* K = 15 

```{r}
#| cache: true
library(boot)
mlr_wage_full <- glm(EXIT_WAGE ~ STARTING_WAGE + AGE_AT_START + SEX_CD +
                    ETHNICITY_CD + EDUCATION_CD + RACE_CD + OCCUP_TYPE + MAJOR_GROUP_NAME, data = clean_data_wage_reg)

set.seed(myseed)
wage_reg_k15 <- cv.glm(clean_data_wage_reg, mlr_wage_full, K = 15)
wage_reg_k15$delta
```

#### Finding the best least squares regression models, using adjusted R-squared, BIC, Mallow’s Cp criterias and Stepwise Variable Selection.

```{r}
library(leaps)
#glimpse(reg_train)
regsub_wage_full <- regsubsets(EXIT_WAGE ~ ., data = reg_train, nvmax = 25, really.big = T)
mlr_wage_full_summary <- summary(regsub_wage_full)
names(mlr_wage_full_summary)
```

##### Best Model based on adjusted R-squared {-}

```{r}
min_adjr2 <- which.max(mlr_wage_full_summary$adjr2)
min_adjr2
adjr2_best <- mlr_wage_full_summary$which[min_adjr2,][mlr_wage_full_summary$which[min_adjr2,] == TRUE][-1]
names(adjr2_best)


```
```{r}
reg_train_r2 = reg_train %>% 
  mutate(
    MAJOR_GROUP_NAME_ArtsHosp_ind = ifelse(MAJOR_GROUP_NAME == "Arts, Hospitality and Sales", 1, 0),
    MAJOR_GROUP_NAME_healthSS_ind = ifelse(MAJOR_GROUP_NAME == "Healthcare and Social Services", 1, 0),
    MAJOR_GROUP_NAME_tradeML_ind = ifelse(MAJOR_GROUP_NAME == "Trades and Manual Labor", 1, 0),
    MAJOR_GROUP_NAME_EducPubServ_ind = ifelse(MAJOR_GROUP_NAME == "Education and Public Services", 1, 0),
     MAJOR_GROUP_NAME_BisManSup_ind = ifelse(MAJOR_GROUP_NAME == "Business, Management and Office Support", 1, 0),
   
    OCCUP_TYPE_Hybrid_ind = ifelse(OCCUP_TYPE == "Hybrid", 1, 0),
    OCCUP_TYPE_TimeBas_ind = ifelse(OCCUP_TYPE == "Time-Based", 1, 0),
    
    EDUCATION_CD_GED_ind = ifelse(EDUCATION_CD == "GED", 1, 0),
    EDUCATION_CD_HSG_ind = ifelse(EDUCATION_CD == "High School graduate (including equivalency)", 1, 0),
    EDUCATION_CD_SomeCol_ind = ifelse(EDUCATION_CD == "Some College or Associate's degree", 1, 0),
    EDUCATION_CD_Bach_ind = ifelse(EDUCATION_CD == "Bachelor's degree", 1, 0),
    
    SEX_Female_ind = ifelse(SEX_CD == "Female", 1, 0),
    
    ETHNICITY_NonHispanic_ind = ifelse(ETHNICITY_CD == "Non-Hispanic or Latino", 1, 0),
    ETHNICITY_Unknown_ind = ifelse(ETHNICITY_CD == "Did Not Self-Identify", 1, 0),

    RACE_black_ind = ifelse(RACE_CD == "Black", 1, 0),
    RACE_asian_ind = ifelse(RACE_CD == "Asian", 1, 0),
    RACE_namer_ind = ifelse(RACE_CD == "Native American", 1, 0),
  ) %>% 
  select(EXIT_WAGE, STARTING_WAGE, AGE_AT_START, contains("_ind"))

glimpse(reg_train_r2)

reg_test_r2 = reg_test %>% 
  mutate(
    MAJOR_GROUP_NAME_ArtsHosp_ind = ifelse(MAJOR_GROUP_NAME == "Arts, Hospitality and Sales", 1, 0),
    MAJOR_GROUP_NAME_healthSS_ind = ifelse(MAJOR_GROUP_NAME == "Healthcare and Social Services", 1, 0),
    MAJOR_GROUP_NAME_tradeML_ind = ifelse(MAJOR_GROUP_NAME == "Trades and Manual Labor", 1, 0),
    MAJOR_GROUP_NAME_EducPubServ_ind = ifelse(MAJOR_GROUP_NAME == "Education and Public Services", 1, 0),
     MAJOR_GROUP_NAME_BisManSup_ind = ifelse(MAJOR_GROUP_NAME == "Business, Management and Office Support", 1, 0),
   
    OCCUP_TYPE_Hybrid_ind = ifelse(OCCUP_TYPE == "Hybrid", 1, 0),
    OCCUP_TYPE_TimeBas_ind = ifelse(OCCUP_TYPE == "Time-Based", 1, 0),
    
    EDUCATION_CD_GED_ind = ifelse(EDUCATION_CD == "GED", 1, 0),
    EDUCATION_CD_HSG_ind = ifelse(EDUCATION_CD == "High School graduate (including equivalency)", 1, 0),
    EDUCATION_CD_SomeCol_ind = ifelse(EDUCATION_CD == "Some College or Associate's degree", 1, 0),
    EDUCATION_CD_Bach_ind = ifelse(EDUCATION_CD == "Bachelor's degree", 1, 0),
    
    SEX_Female_ind = ifelse(SEX_CD == "Female", 1, 0),
    
    ETHNICITY_NonHispanic_ind = ifelse(ETHNICITY_CD == "Non-Hispanic or Latino", 1, 0),
    ETHNICITY_Unknown_ind = ifelse(ETHNICITY_CD == "Did Not Self-Identify", 1, 0),

    RACE_black_ind = ifelse(RACE_CD == "Black", 1, 0),
    RACE_asian_ind = ifelse(RACE_CD == "Asian", 1, 0),
    RACE_namer_ind = ifelse(RACE_CD == "Native American", 1, 0),
  ) %>% 
  select(EXIT_WAGE, STARTING_WAGE, AGE_AT_START, contains("_ind"))

glimpse(reg_test_r2)
```



```{r}
mlr_wage_adjr2_best <- lm(EXIT_WAGE ~ ., data = reg_train_r2)
summary(mlr_wage_adjr2_best)
adjr2_predict = predict(mlr_wage_adjr2_best, reg_test_r2)
mean((reg_test_r2$EXIT_WAGE - adjr2_predict)^2)
```

##### Best Model based on BIC: {-}

```{r}
round(mlr_wage_full_summary$bic, 0)
max_BIC <- which.max(mlr_wage_full_summary$bic) # The worst by BIC
min_BIC <- which.min(mlr_wage_full_summary$bic) # The best by BIC
min_BIC
mlr_wage_full_summary$which[min_BIC,][mlr_wage_full_summary$which[min_BIC,] == TRUE][-1]
```

```{r}
mlr_wage_bic_best <- lm(EXIT_WAGE ~ . -MAJOR_GROUP_NAME_healthSS_ind - MAJOR_GROUP_NAME_ArtsHosp_ind -MAJOR_GROUP_NAME_BisManSup_ind
                        -OCCUP_TYPE_Hybrid_ind -OCCUP_TYPE_TimeBas_ind -EDUCATION_CD_Bach_ind, data = reg_train_r2)
summary(mlr_wage_bic_best)
bic_predict = predict(mlr_wage_bic_best, reg_test_r2)
mean((reg_test_r2$EXIT_WAGE - bic_predict)^2)
```

##### Best Model based on Mallow’s Cp {-}

```{r}
round(mlr_wage_full_summary$cp, 2)
max_Cp <- which.max(abs(mlr_wage_full_summary$cp - 1:17)) # The worst by Cp - underfit so high bias.
max_Cp <- which.min(abs(mlr_wage_full_summary$cp - 1:17)) # The best by Cp
max_Cp
names(mlr_wage_full_summary$which[max_Cp,][mlr_wage_full_summary$which[max_Cp,] == TRUE])[-1]
```

```{r}
mlr_wage_cp_best <- lm(EXIT_WAGE ~ . -MAJOR_GROUP_NAME_healthSS_ind - MAJOR_GROUP_NAME_ArtsHosp_ind -MAJOR_GROUP_NAME_BisManSup_ind
                       -EDUCATION_CD_Bach_ind, 
                       data = reg_train_r2)
summary(mlr_wage_cp_best)
cp_predict = predict(mlr_wage_cp_best, reg_test_r2)
mean((reg_test_r2$EXIT_WAGE - cp_predict)^2)
```


##### Best Model based on Stepwise Forward and Backward Selection {-}

```{r}
reg_null <- lm(EXIT_WAGE ~ 1, data = reg_train)
```



```{r}
step_forward <- step(reg_null,
                     scope = list(lower = reg_null, upper = mlr_wage),
                     direction = "forward", 
                     trace = 0)
summary(step_forward)
```


```{r}
step_outb <- step(mlr_wage, 
                 scope = list(lower = reg_null, upper = mlr_wage),
                 method = "backward", 
                 trace = 0)
summary(step_outb)
```

```{r}
AIC(mlr_wage, step_outb, step_forward)
```



### Ridge Regression

```{r}
x_matrix <- model.matrix(mlr_wage_full)[, -1]
dim(x_matrix)

library(glmnet)

set.seed(myseed)
rr_cv <- cv.glmnet(x_matrix[w_Z,], y = reg_train$EXIT_WAGE, alpha = 0)
rr_cv
```

```{r}
plot(rr_cv)
```

```{r}
coef(rr_cv)
```

```{r}
Yhat_ridge <- predict(rr_cv, newx = x_matrix[-w_Z,], s = "lambda.min")
mean((reg_test$EXIT_WAGE - Yhat_ridge)^2) |> round(3)
```


### LASSO Regression

```{r}
set.seed(myseed)
lr_cv <- cv.glmnet(x_matrix[w_Z,], y = reg_train$EXIT_WAGE, alpha = 1)
lr_cv
```

```{r}
coef(lr_cv) 
```

```{r}
library(vip)
vip::vip(lr_cv)
```

```{r}
Yhat_lasso <- predict(lr_cv, newx = x_matrix[-w_Z,], s = "lambda.min")
mean((reg_test$EXIT_WAGE - Yhat_lasso)^2) |> round(3)
```

```{r}
Yhat_lasso <- predict(lr_cv, newx = x_matrix[-w_Z,], s = "lambda.1se")
mean((reg_test$EXIT_WAGE - Yhat_lasso)^2) |> round(3)
```


### PCR with K = 10 cross-validation


```{r}
library(pls)
set.seed(myseed)
pcrv_fit <- pcr(EXIT_WAGE ~ STARTING_WAGE + AGE_AT_START + SEX_CD + ETHNICITY_CD + RACE_CD + OCCUP_TYPE,
                data = reg_train, scale = TRUE, validation = "CV", segments = 10)
summary(pcrv_fit)
```

```{r}
vip::vip(pcrv_fit)
```

```{r}
opt_M <- which.min(pcrv_fit$validation$PRESS)
opt_M
```

```{r}
pcr_pred <- predict(pcrv_fit, reg_test, ncomp = opt_M)
test_MSE <- mean((pcr_pred - reg_test$EXIT_WAGE)^2)
test_MSE
```


### PLS with K = 10 cross-validation

```{r}
set.seed(myseed)
plsrv_fit <- plsr(EXIT_WAGE ~ STARTING_WAGE + AGE_AT_START + SEX_CD + ETHNICITY_CD + RACE_CD + OCCUP_TYPE, 
                  data = reg_train, scale = TRUE, validation = "CV", segments = 10)
summary(plsrv_fit)
```


```{r}
vip::vip(plsrv_fit)
```


```{r}
opt_M <- which.min(plsrv_fit$validation$PRESS)
opt_M
```


```{r}
plsr_pred <- predict(plsrv_fit, reg_test, ncomp = opt_M)
test_MSE <- mean((plsr_pred - reg_test$EXIT_WAGE)^2)
test_MSE
```


### Regression Trees

```{r}
library(tree)
glimpse(reg_train)
tree_wage  <- tree(EXIT_WAGE ~ STARTING_WAGE + AGE_AT_START + OCCUP_TYPE + EDUCATION_CD + 
                   SEX_CD + ETHNICITY_CD + RACE_CD + MAJOR_GROUP_NAME, 
                   data = reg_train)
tree_wage
```


```{r}
summary(tree_wage)
```


```{r}
plot(tree_wage)
text(tree_wage)
```


```{r}
tree_wage_pred = predict(tree_wage, newdata = reg_test)
mean((tree_wage_pred - reg_test$EXIT_WAGE)^2)
```

#### Optimal tree based on cross-validation mean squared error

```{r}
set.seed(myseed)
treem_cvm <- cv.tree(tree_wage)
treem_cvm$dev |> round(0)

treem_cvm$size[which.min(treem_cvm$dev)]
```

```{r}
plot(treem_cvm)
```

The tree has 5 terminal nodes.

The lowest total deviance is 3983000 and we can use it to estimate the accuracy of the tree: 3983000 / 44180 = 13.28. This value is a more reliable estimate of performance in predicting mpg with these four predictors.


```{r}
MSE = 2784473 / (30925 - 5)
MSE
```

#### Random Forest

* Results using the default model and parameters yield by the `randomForest` package.

```{r}
#| cache: true
library(randomForest)
set.seed(myseed)
wage_rf <- randomForest(EXIT_WAGE ~ STARTING_WAGE + AGE_AT_START + OCCUP_TYPE + EDUCATION_CD + 
                        SEX_CD + ETHNICITY_CD + RACE_CD + MAJOR_GROUP_NAME, 
                        data = reg_train, importance = TRUE)
wage_rf
```


```{r}
Yhat_rf <-  predict(wage_rf, newdata = reg_test)
mean((Yhat_rf - reg_test$EXIT_WAGE)^2)
```

The prediction MSE is 72.51022 which is the lowest computed so far.


```{r}
which.min(wage_rf$mse)
```


```{r}
#importance(wage_rf)
importance(wage_rf)[order(importance(wage_rf)[, 1], decreasing = TRUE), ] |> round(0)
```

```{r}
# #| cache: true
# set.seed(myseed)
# tuneRF(x = reg_train, y = reg_train$EXIT_WAGE)
```



##### Random Forest Model Tuning

I'm using the `ranger` package because it is much faster and more accurate than the original `randomForest` package, and it's more suitable for large datasets.

This prints: Cross-validated RMSE, Best hyperparameters (Cross-validated hyperparameter tuning), Model structure.

```{r}
#| cache: true

library(caret)
library(ranger)

set.seed(myseed)

tc <- trainControl(method = "cv", number = 10)

rf_grid <- expand.grid(
  mtry = c(8),
  splitrule = "variance",
  min.node.size = c(5, 10, 20)
)

rf_model <- train(
  EXIT_WAGE ~ .,
  data = reg_train,
  method = "ranger",
  trControl = tc,
  tuneGrid = rf_grid,
  importance = "impurity",
  num.trees = 500,
  metric = "RMSE"
)

rf_model
```


# Compute Test RMSE

```{r}
rf_pred <- predict(rf_model, newdata = reg_test)

mean((rf_pred - reg_test$EXIT_WAGE)^2)
rf_rmse <- sqrt(mean((rf_pred - reg_test$EXIT_WAGE)^2))
rf_rmse
```

# Check variable importance

```{r}
varImp(rf_model)
plot(varImp(rf_model))
```


* We can see that STARTING_WAGE is dominating importance.

* AGE_AT_START, MAJOR_GROUP_NAME, several education/race variables, OCCUP_TYPE are also important, with some contributing small amounts each for the model.

